---
title: "associations-between-outcomes-analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data
## Test data
test_data_long <- readRDS(file = "data/derivedData/test_data_long")

test_data_wide <- readRDS(file = "data/derivedData/test_data_wide")


```



# Explorative analysis

To explore whether within-individual response variability in one outcome (e.g. Hbmass) was associated the within-individual response variability in another outcome (e.g. V̇O2max), each participant’s between-period difference in change scores (∆ ET2 - ∆ ET1) was calculated for selected outcomes, and linear regression models were fitted for each outcome pair of interest (R stats package) on the form:
outcome A ~ outcome B
The coefficient of determination (R2) was retrieved from each model and bootstrapped 95% confidence intervals for R² were obtained using nonparametric bootstrap resampling with 1000 resamples. Pearson´s correlation coefficients (r) with 95% CIs (R stats package) were also used to examine the associations.


```{r}
#| label: Associations within-individual variability across outcomes

diff_selected_outcomes <- test_data_long %>% 
  select(id, period, variable, p.change.pre.post) %>% 
  filter(variable %in% c("cs_umol.g.min", "had_umol.g.min")) %>% 
  pivot_wider(names_from = period, values_from = p.change.pre.post) %>% 
  mutate(diff.p1.p2 = (`2` - `1`)) %>% 
  select(id, variable, diff.p1.p2) %>% 
  drop_na() 


diff_outcomes <- test_data_long %>% 
  select(id, period, variable, change.pre.post) %>% 
  filter(variable %in% c("watt.15tt", "watt.max", "vo2.max", "p.vo2.15tt", "ge.submax", "hbmass", "bv", "cd", "cf")) %>% 
  pivot_wider(names_from = period, values_from = change.pre.post) %>% 
  mutate(diff.p1.p2 = (`2` - `1`)) %>%
  select(id, variable, diff.p1.p2) %>%
  full_join(diff_selected_outcomes, by = c("id", "variable", "diff.p1.p2"))


# Define physiological and performance outcomes
phys_outcomes  <- c("hbmass", "bv", "cd", "cf", "cs_umol.g.min", "had_umol.g.min")
perf_outcomes  <- c("watt.15tt", "watt.max", "vo2.max")

# Function to compute R² and Pearson r with 95% CI using cor.test
bootstrap_r2_r <- function(data, outcome, predictor, nboot = 1000, seed = 123) {
  
  set.seed(seed)
  
  # Subset and reshape data
  r_data <- data %>%
    filter(variable %in% c(outcome, predictor)) %>%
    pivot_wider(names_from = variable, values_from = diff.p1.p2) %>%
    drop_na()
  
  # Skip if too few data points
  if (nrow(r_data) < 5) {
    return(tibble(
      predictor   = predictor,
      outcome     = outcome,
      r_squared   = NA_real_,
      r2_low      = NA_real_,
      r2_high     = NA_real_,
      pearsons_r  = NA_real_,
      r_low       = NA_real_,
      r_high      = NA_real_
    ))
  }
  
  # Linear regression R²
  fit <- lm(r_data[[outcome]] ~ r_data[[predictor]])
  r2_point <- summary(fit)$r.squared
  
  # Bootstrap R²
  r2_boot <- replicate(nboot, {
    idx <- sample(seq_len(nrow(r_data)), replace = TRUE)
    fit_b <- lm(r_data[[outcome]][idx] ~ r_data[[predictor]][idx])
    summary(fit_b)$r.squared
  })
  r2_ci <- quantile(r2_boot, probs = c(0.025, 0.975), na.rm = TRUE)
  
  # Pearson r and 95% CI using cor.test
  cor_res <- cor.test(r_data[[outcome]], r_data[[predictor]], method = "pearson")
  
  tibble(
    predictor   = predictor,
    outcome     = outcome,
    r_squared   = r2_point,
    r2_low      = r2_ci[1],
    r2_high     = r2_ci[2],
    pearsons_r  = unname(cor_res$estimate),
    r_low       = cor_res$conf.int[1],
    r_high      = cor_res$conf.int[2]
  )
}

# Compute all combinations of performance and physiological outcomes
results <- purrr::cross_df(list(
  outcome   = perf_outcomes,
  predictor = phys_outcomes
)) %>%
  pmap_dfr(~ bootstrap_r2_r(
    data      = diff_outcomes,
    outcome   = ..1,
    predictor = ..2,
    nboot     = 1000
  )) %>%
  mutate(
    r_squared   = round(r_squared, 2),
    r2_low      = round(r2_low, 2),
    r2_high     = round(r2_high, 2),
    pearsons_r  = round(pearsons_r, 2),
    r_low       = round(r_low, 2),
    r_high      = round(r_high, 2)
  )

# Format for presentation
results_summary <- results %>%
  mutate(
    `R2 [95% CI]` = sprintf("%.2f [%.2f, %.2f]", r_squared, r2_low, r2_high),
    `r [95% CI]`  = sprintf("%.2f [%.2f, %.2f]", pearsons_r, r_low, r_high)
  ) %>%
  select(outcome, predictor, `R2 [95% CI]`, `r [95% CI]`)

print(results_summary, width = Inf, n = Inf)



```

