---
title: "associations-between-outcomes-analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data
## Test data
test_data_long <- readRDS(file = "data/derivedData/test_data_long")

test_data_wide <- readRDS(file = "data/derivedData/test_data_wide")


```

# Explorative analysis

### 1. Associations within-individual variability across outcomes

### 2. Effect of age on the association between training adaptations in period 1 and 2

### 3. Effect of sex on the association between training adaptations in period 1 and 2


### 1. Associations within-individual variability across outcomes

To explore whether within-individual response variability in one outcome (e.g. Hbmass) was associated the within-individual response variability in another outcome (e.g. V̇O2max), each participant’s between-period difference in change scores (∆ ET2 - ∆ ET1) was calculated for selected outcomes, and linear regression models were fitted for each outcome pair of interest (R stats package) on the form: outcome A \~ outcome B The coefficient of determination (R2) was retrieved from each model and bootstrapped 95% confidence intervals for R² were obtained using nonparametric bootstrap resampling with 1000 resamples. Pearson´s correlation coefficients (r) with 95% CIs (R stats package) were also used to examine the associations.

```{r}
#| label: Associations within-individual variability across outcomes

diff_selected_outcomes <- test_data_long %>% 
  select(id, period, variable, p.change.pre.post) %>% 
  filter(variable %in% c("cs_umol.g.min", "had_umol.g.min")) %>% 
  pivot_wider(names_from = period, values_from = p.change.pre.post) %>% 
  mutate(diff.p1.p2 = (`2` - `1`)) %>% 
  select(id, variable, diff.p1.p2) %>% 
  drop_na() 


diff_outcomes <- test_data_long %>% 
  select(id, period, variable, change.pre.post) %>% 
  filter(variable %in% c("watt.15tt", "watt.max", "vo2.max", "p.vo2.15tt", "ge.submax", "hbmass", "bv", "cd", "cf")) %>% 
  pivot_wider(names_from = period, values_from = change.pre.post) %>% 
  mutate(diff.p1.p2 = (`2` - `1`)) %>%
  select(id, variable, diff.p1.p2) %>%
  full_join(diff_selected_outcomes, by = c("id", "variable", "diff.p1.p2"))


# Define physiological and performance outcomes
phys_outcomes  <- c("hbmass", "bv", "cd", "cf", "cs_umol.g.min", "had_umol.g.min")
perf_outcomes  <- c("watt.15tt", "watt.max", "vo2.max")

# Function to compute R² and Pearson r with 95% CI using cor.test
bootstrap_r2_r <- function(data, outcome, predictor, nboot = 1000, seed = 123) {
  
  set.seed(seed)
  
  # Subset and reshape data
  r_data <- data %>%
    filter(variable %in% c(outcome, predictor)) %>%
    pivot_wider(names_from = variable, values_from = diff.p1.p2) %>%
    drop_na()
  
  # Skip if too few data points
  if (nrow(r_data) < 5) {
    return(tibble(
      predictor   = predictor,
      outcome     = outcome,
      r_squared   = NA_real_,
      r2_low      = NA_real_,
      r2_high     = NA_real_,
      pearsons_r  = NA_real_,
      r_low       = NA_real_,
      r_high      = NA_real_
    ))
  }
  
  # Linear regression R²
  fit <- lm(r_data[[outcome]] ~ r_data[[predictor]])
  r2_point <- summary(fit)$r.squared
  
  # Bootstrap R²
  r2_boot <- replicate(nboot, {
    idx <- sample(seq_len(nrow(r_data)), replace = TRUE)
    fit_b <- lm(r_data[[outcome]][idx] ~ r_data[[predictor]][idx])
    summary(fit_b)$r.squared
  })
  r2_ci <- quantile(r2_boot, probs = c(0.025, 0.975), na.rm = TRUE)
  
  # Pearson r and 95% CI using cor.test
  cor_res <- cor.test(r_data[[outcome]], r_data[[predictor]], method = "pearson")
  
  tibble(
    predictor   = predictor,
    outcome     = outcome,
    r_squared   = r2_point,
    r2_low      = r2_ci[1],
    r2_high     = r2_ci[2],
    pearsons_r  = unname(cor_res$estimate),
    r_low       = cor_res$conf.int[1],
    r_high      = cor_res$conf.int[2]
  )
}

# Compute all combinations of performance and physiological outcomes
results <- purrr::cross_df(list(
  outcome   = perf_outcomes,
  predictor = phys_outcomes
)) %>%
  pmap_dfr(~ bootstrap_r2_r(
    data      = diff_outcomes,
    outcome   = ..1,
    predictor = ..2,
    nboot     = 1000
  )) %>%
  mutate(
    r_squared   = round(r_squared, 2),
    r2_low      = round(r2_low, 2),
    r2_high     = round(r2_high, 2),
    pearsons_r  = round(pearsons_r, 2),
    r_low       = round(r_low, 2),
    r_high      = round(r_high, 2)
  )

# Format for presentation
results_summary <- results %>%
  mutate(
    `R2 [95% CI]` = sprintf("%.2f [%.2f, %.2f]", r_squared, r2_low, r2_high),
    `r [95% CI]`  = sprintf("%.2f [%.2f, %.2f]", pearsons_r, r_low, r_high)
  ) %>%
  select(outcome, predictor, `R2 [95% CI]`, `r [95% CI]`)

print(results_summary, width = Inf, n = Inf)



```



### 2. Effect of age on the association between training adaptations in period 1 and 2

```{r}
#| label: Effect of age on the association between training adaptations in period 1 and 2

age_data <- test_data_wide %>% 
  filter(period == "1") %>% 
  filter(timepoint == "pre") %>% 
  select(id, age) %>% print()

# Absolute changes

age_responses_absolute_data <- test_data_long %>% 
  inner_join(age_data, join_by(id)) %>% 
  select(id, period, age, change.pre.post) %>%
  pivot_wider(names_from = period,
              values_from = change.pre.post) %>% 
  rename(change.p1 = `1`) %>% 
  rename(change.p2 = `2`) 


outcomes <- c("watt.15tt", "watt.15tt.bm", "watt.15tt.lbm", 
              "watt.max", "watt.max.bm", "watt.max.lbm",
              "vo2.max", "vo2.max.bm", "vo2.max.lbm")


results_absolute <- map_dfr(outcomes, function(outcome) {
  
  model <- lm(
    change.p2 ~ change.p1 * age, data = age_responses_absolute_data %>% 
      filter(variable == outcome))
  
  tidy(model) %>% 
    filter(term == "change.p1:age") %>% 
    mutate(outcome = outcome) %>% 
    select(outcome, estimate, p.value)
})

print(results_absolute)



# Percentage changes

age_responses_percentage_data <- test_data_long %>% 
  inner_join(age_data, join_by(id)) %>% 
  select(id, period, age, p.change.pre.post) %>%
  pivot_wider(names_from = period,
              values_from = p.change.pre.post) %>% 
  rename(change.p1 = `1`) %>% 
  rename(change.p2 = `2`)


outcomes <- c("watt.15tt", "watt.max", "vo2.max")


results_percentage <- map_dfr(outcomes, function(outcome) {
  
  model <- lm(
    change.p2 ~ change.p1 * age, data = age_responses_percentage_data %>% 
      filter(variable == outcome))
  
  tidy(model) %>% 
    filter(term == "change.p1:age") %>% 
    mutate(outcome = outcome) %>% 
    select(outcome, estimate, p.value)
})

print(results_percentage)

```


### 3. Effect of sex on the association between training adaptations in period 1 and 2

```{r}
#| label: Effect of sex on the association between training adaptations in period 1 and 2


# Absolute changes

sex_responses_absolute_data <- test_data_long %>% 
  select(id, period, sex, change.pre.post) %>% 
  pivot_wider(names_from = period,
              values_from = change.pre.post) %>% 
  rename(change.p1 = `1`) %>% 
  rename(change.p2 = `2`) 


outcomes <- c("watt.15tt", "watt.15tt.bm", "watt.15tt.lbm", 
              "watt.max", "watt.max.bm", "watt.max.lbm",
              "vo2.max", "vo2.max.bm", "vo2.max.lbm")


results_absolute <- map_dfr(outcomes, function(outcome) {
  
  model <- lm(
    change.p2 ~ change.p1 * sex, data = sex_responses_absolute_data %>% 
      filter(variable == outcome))
  
  tidy(model) %>% 
    filter(term == "change.p1:sexm") %>% 
    mutate(outcome = outcome) %>% 
    select(outcome, estimate, p.value)
})

print(results_absolute)


# Percentage changes

sex_responses_percentage_data <- test_data_long %>% 
  select(id, period, sex, p.change.pre.post) %>% 
  pivot_wider(names_from = period,
              values_from = p.change.pre.post) %>% 
  rename(change.p1 = `1`) %>% 
  rename(change.p2 = `2`) 


outcomes <- c("watt.15tt", "watt.max", "vo2.max")


results_absolute <- map_dfr(outcomes, function(outcome) {
  
  model <- lm(
    change.p2 ~ change.p1 * sex, data = sex_responses_percentage_data %>% 
      filter(variable == outcome))
  
  tidy(model) %>% 
    filter(term == "change.p1:sexm") %>% 
    mutate(outcome = outcome) %>% 
    select(outcome, estimate, p.value)
})

print(results_absolute)



```



### 4. Reproducibility of individual training responses in females and males

```{r}
#| label: Reproducibility of individual training responses in females and males


# Females

## Absolute changes
test_data_long_females <- test_data_long %>% 
  filter(sex == "f") 

# List of variables you want to process
outcomes <- c("watt.15tt", "watt.15tt.bm", "watt.15tt.lbm", 
               "watt.max", "watt.max.bm", "watt.max.lbm", 
               "vo2.max","vo2.max.bm", "vo2.max.lbm") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs

analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long_females,       # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary_absolute_females <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary_absolute_females, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, change.pre.post) %>%
    filter(!is.na(change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long_females, var)
}

# Print the results data frame
print(results)





## % changes of absolute values

# List of variables you want to process
outcomes <- c("watt.15tt", "watt.max", "vo2.max") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(p.change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    p.change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long_females,             # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary_absolute_females <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary_absolute_females, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, p.change.pre.post) %>%
    filter(!is.na(p.change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = p.change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long_females, var)
}

# Print the results data frame
print(results)








## Males


## Absolute changes
test_data_long_males <- test_data_long %>% 
  filter(sex == "m") 

# List of variables you want to process
outcomes <- c("watt.15tt", "watt.15tt.bm", "watt.15tt.lbm", 
               "watt.max", "watt.max.bm", "watt.max.lbm", 
               "vo2.max","vo2.max.bm", "vo2.max.lbm") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs

analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long_males,       # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary_absolute_males <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary_absolute_males, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, change.pre.post) %>%
    filter(!is.na(change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long_males, var)
}

# Print the results data frame
print(results)





## % changes of absolute values

# List of variables you want to process
outcomes <- c("watt.15tt", "watt.max", "vo2.max") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(p.change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    p.change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long_males,             # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary_percentage_males <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary_percentage_males, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, p.change.pre.post) %>%
    filter(!is.na(p.change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = p.change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long_males, var)
}

# Print the results data frame
print(results)



```




