---
title: "data-preparation"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# RepeatStudy-repeated-endurance-training-paper

In this repository all data preparation and analysis is done for the following manuscript:

Limited individual reproducibility of physiological adaptations to repeated endurance exercise training

Authors: Ingvill Urianstad Odden, Håvard Hamarsland, Tomas Urianstad Odden, Daniel Hammarström, Anne Mette Rustaden, Lise Koll, Marita Hanestadhaugen, Steen Larsen, Stian Ellefsen, Joar Hansen, Håvard Nygaard, Carsten Lundby, Knut Sindre Mølmen

## Data preparation

In this file, all data is loaded from the RepeatData package (not publicly available) and prepared for analyses.

### Loading needed packages for data preparation and analysis

```{r, load packages}
#| label: Packages 

# Loading needed packages
library(tidyverse)
library(repeatData)
library(readxl)

```

#### Periods:

1 and 2 refers to training period 1 and 2 (ET-1 and ET-2 in manuscript)

#### Timepoint:

Testing timepoint, differs between training periods 1 and 2.

##### Period 1

Period 1 timepoint 1 = participants performed submaximal testing and incremental VO2max testing. This was the familarization testing (hence not used in analysis)

Period 1 timepoint 2 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. Submaximal test had different absolute power output than at timepoint 3 and 7 (hence not used in analysis). 15-min test was familarization (hence not used in analysis). Incremental VO2max testing was regarded participants´ PRE 1 test.

Period 1 timepoint 3 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. Submaximal test and 15-min test regarded as participants´ single PRE test. Incremental VO2max testing was regarded participants´ PRE 2 test.

Period 1 timepoint 7 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. All tests regarded participants´ single POST test.

##### Period 2

Period 2 timepoint 1 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. Submaximal test had different absolute power output than at timepoint 2 and 3 (hence not used in analysis). Incremental VO2max testing and 15-min test were regarded participants´ PRE 1 test.

Period 2 timepoint 2 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. Submaximal test regarded as participants´ single PRE test. Incremental VO2max testing and 15-min test were regarded participants´ PRE 2 test.

Period 2 timepoint 3 = participants performed submaximal testing, incremental VO2max testing, and 15-min performance test. All tests regarded participants´ single POST test.

### Cycling data:

Preparing data from the cycling data set, i.e. the exercise testing data.

Duplicate pre-tests was averaged to a single PRE value. Duplicate tests were also used to calculate TE and %CV.

```{r, cycling data}
#| label: Cycling data


# Submaximal cycling data

filtered_submax_data <- cycling.data %>%
  filter(period %in% c(1, 2)) %>% 
  filter(group == "intervention") %>% 
  filter(test == "sub_40") %>% # Only the second of the two submaximal bouts (25 and 40% of MMPOVO2max) is used in data analysis
  
  filter((period == 1 & timepoint %in% c(3, 7)) | 
          # Timepoint 3 =  pre2, timepoint 7 = post 
          # Only pre2 is used because only this pre-test had the same absolute power output at the submaximal bout as post-tests)
         (period == 2 & timepoint %in% c(2, 3))) %>% 
          # Timepoint 2 = pre2, timepoint 3 = post
          # Only pre2 is used because only this pre-test had the same absolute power output at the submaximal bout as post-tests)
 
   mutate(timepoint = case_when(
    timepoint == 3 & period == 1 ~ "pre2",
    timepoint == 7 & period == 1 ~ "post",
    timepoint == 2 & period == 2 ~ "pre2",
    timepoint == 3 & period == 2 ~ "post")) %>% 
  
  select(id, group, period, timepoint, watt.submax = watt, 
         vo2.submax = vo2, rpe.submax = rpe, lac.submax = lac, hr.submax = hr, 
         ve.submax = ve, rer.submax = rer, ge.submax = GE) %>% 
  filter(!(id %in% c("49")))
  # Removing participant (id49) because missing pre2 in period 1. Can hence not be used in analysis of submaximal tests
  
   

# Maximal incremental cycling data

filtered_max_data <- cycling.data %>%
  filter(period %in% c(1, 2)) %>%
  filter(group == "intervention") %>% 
  filter(test == "max") %>% 
  
  filter((period == 1 & timepoint %in% c(2, 3, 7)) | 
          # Timepoints 2 and 3 = pre1 and pre2, timepoint 7 = post
         (period == 2 & timepoint %in% c(1, 2, 3))) %>% 
          # Timepoints 1 and 2 = pre1 and pre2, timepoint 3 = post
  
  mutate(timepoint = case_when(
    timepoint == 2 & period == 1 ~ "pre1",
    timepoint == 3 & period == 1 ~ "pre2",
    timepoint == 7 & period == 1 ~ "post",
    timepoint == 1 & period == 2 ~ "pre1",
    timepoint == 2 & period == 2 ~ "pre2",
    timepoint == 3 & period == 2 ~ "post")) %>% 
  
  select(id, group, period, timepoint, sex, age, height, weight, 
         watt.tresh = threshold, watt.max = watt, vo2.max = vo2, 
         rpe.max = rpe, lac.max = lac, hr.max = hr, ve.max = ve, rer.max = rer) 
  # Id49 have data from pre1, but not pre2. Pre 1 is hence used in analysis



# Missing performance 15-min cycling  data for id49 at pre2 period 1, have to use familarization data instead

id49per_data <- cycling.data %>%
  filter(period %in% c(1, 2)) %>% 
  filter(group == "intervention") %>% 
  filter(test == "per") %>% 
  filter(period == 1, timepoint %in% c(2), id %in% c(49)) %>%  
  mutate(timepoint = "pre2") 
  # Id49 did not perform the 15-min test at timepoint 3 (pre2) period 1.
  # As well-performed test at timepoint 2 (familarization), we have chosen to use this test as timepoint 3 (pre2) in period 1 in data analysis



# Performance 15-min cycling data

filtered_per_data <- cycling.data %>% 
  filter(period %in% c(1, 2)) %>% 
  filter(group == "intervention") %>% 
  filter(test == "per") %>% 
      
  filter((period == 1 & timepoint %in% c(3, 7)) | 
          # Timepoint 3 = pre2, timepoint 7 = post
         (period == 2 & timepoint %in% c(1, 2, 3))) %>% 
          # Timepoints 1 and 2 = pre1 and pre2, timepoint 3 = post
  
  mutate(timepoint = case_when(
    timepoint == 3 & period == 1 ~ "pre2", # Only one 15 min test at pre2 (not pre1) in period 1
    timepoint == 7 & period == 1 ~ "post",
    timepoint == 1 & period == 2 ~ "pre1",
    timepoint == 2 & period == 2 ~ "pre2",
    timepoint == 3 & period == 2 ~ "post")) %>% 
  
  full_join(id49per_data) %>% # Adding data for id49
  
  select(id, group, period, timepoint, watt.15tt = watt,
         vo2.15tt = vo2, rpe.15tt = rpe, lac.15tt = lac, hr.15tt = hr, 
         ve.15tt = ve, rer.15tt = rer, p.vo2.15tt = rel.vo2)



# Combining filtered cycling data

all_timepoints_cycling_data <- filtered_max_data %>% 
  full_join(filtered_submax_data, by = c("id", "period", "timepoint", "group")) %>%
  full_join(filtered_per_data, by = c("id", "period", "timepoint", "group"))
  
cycling_data <- all_timepoints_cycling_data %>% 
  filter(timepoint %in% c("pre1", "pre2")) %>%
  # Calculate the mean of pre1 and pre2 and rename to pre
  
  group_by(id, group, period) %>%
  summarise(                
    group = first(group),
    period = first(period),
    timepoint = "pre",
    sex = first(sex),
    across(where(is.numeric), mean, na.rm = TRUE),
    .groups = "drop") %>% 
  
  # Combine with post
  bind_rows(
    all_timepoints_cycling_data %>%
      filter(timepoint %in% c("post")) %>%
      select(id, period, timepoint, sex, everything())) %>%
  arrange(id, period, timepoint) %>%
  select(-group)


 # Save as RDS for later use 
#saveRDS(cycling_data, file = "data/derivedData/cycling_data")


```

#### All other data was collected at one pre-test and one post-test in period 1 and 2, respectively.

### Body composition data:

Preparing dual-energy X-ray absorptiometry (DXA) scan data.

```{r, dxa}
#| label: body composition

body_composition_data <- dxa.data %>% 
  filter(period %in% c(1, 2)) %>%
  
  mutate(id = as.factor(id)) %>% # Convert id to factor
  mutate(period = as.character(period)) %>% # Convert period to character
  
  select(id, period, timepoint, lean.mass.whole.body) 
    # Selecting which variables to use in the present analyses
  

# Save as RDS for later use 
#saveRDS(body_composition_data, file = "data/derivedData/body_composition_data")

```

### Immunohistochemistry:

```{r, ihc}
#| label: Immunohistochemistry



immuno_data <- immuno.data %>% 
  filter(period %in% c(1, 2)) %>%
  # Participants missing data is already filtered out in the data. 
  # Data from 13 participants missing (due to missing data at least one timepoint)
  
  select(id, period, timepoint, fiber.area, cd, cf) %>% 
  # Selecting the variables of interest
  
  mutate(fiber.area = as.numeric(fiber.area), # Convert variables to numeric
         cd = as.numeric(cd),
         cf = as.numeric(cf)) %>%
  mutate(id = as.factor(id)) %>% # Convert id to factor
  mutate(period = as.character(period)) # Convert period to character

# Save as RDS for later use 
#saveRDS(immuno_data, file = "data/derivedData/immuno_data")


```

### CS and HAD activities data:

```{r, cs/had}
#| label: CS/HAD 

cs_had_data <- cs_had.data %>% 
  filter(period %in% c(1, 2)) %>%
  
  filter(!id %in% c(33, 49)) %>% 
  # Filtering out participants who did not have measurements in both period 1 and 2
  
  mutate(id = as.numeric(as.character(id)), # Convert id to numeric
         id = as.factor(id))  # Convert id back to a factor
  
# Save as RDS for later use 
#saveRDS(cs_had_data, file = "data/derivedData/cs_had_data")

```

### Hematological data:

```{r, blood}
#| label: Blood

# Creating hematological_data

hematological_data <- hb.data %>% 
  filter(period %in% c(1, 2)) %>% 
  filter(group == "intervention") %>% 
  select(-weight) %>% # Also added in this data set, but self reported, so will use the measured weight from exercise test in cycling data to claculate relative hemoglobin mass
  select(-hbmass.kg) %>%
  select(-group) 

# Save as RDS for later use 
#saveRDS(hematological_data, file = "data/derivedData/hematological_data")


```

### Ferritin data

```{r, Fe}
#| label: Ferritin 

ferritin_data <- iron_status.data %>% 
  filter(period %in% c(1, 2)) %>%
  
  mutate(id = as.factor(id)) %>% # Convert id to factor
  mutate(period = as.character(period)) %>% # Convert period to character
  mutate(ferritin.level = as.numeric(iron_status)) %>% # Convert variable to numeric
  select(-iron_status) %>% 
  
  filter(!(id %in% c("50", "55"))) %>% 
  # Due to missing blood samples
  
  
  mutate(id = as.numeric(as.character(id)), # Convert id to numeric
         id = as.factor(id)) %>% view() # Convert id back to a factor


# Save as RDS for later use 
#saveRDS(ferritin_data, file = "data/derivedData/ferritin_data")

```


### Other blood data

```{r, LDL, HDL, triglyserides, glucose, crp, c peptide}

blood_markers_data <- blood_variables.data %>% 
  filter(group == "intervention") %>% 
  select(-group) %>% 
  select(-comment) %>% 
  filter(timepoint %in% c(1, 2, 3, 4)) %>% 
  # Timepoints 1 and 2 = pre and post period 1
  # Timepoints 3 and 4 = pre and post period 3

  mutate(
    period = case_when(
      timepoint %in% c(1, 2) ~ 1,
      timepoint %in% c(3, 4) ~ 2), # Make period column
    
    timepoint = case_when(
      timepoint %in% c(1, 3) ~ "pre",
      timepoint %in% c(2, 4) ~ "post")) %>% # Changing name of timepoints
  
  #filter(period == "1") %>% 
  #filter(timepoint == "pre") %>% 
  
  filter(!(id %in% c("50", "55"))) %>% 
  # Due to missing blood samples
  
  mutate(period = as.factor(period)) %>% # Convert period to factor
  select(id, period, timepoint, c.peptid) %>%  
  view()



```

### Combined test data

Combining all test data. Also scaling values to total body mass (bm) and lean body mass (lbm)

```{r, full data}
#| label: Data combined 

# Data in wide format
test_data_wide <- cycling_data %>% 
  full_join(body_composition_data, by = c("id", "period", "timepoint")) %>% 
  full_join(immuno_data, by = c("id", "period", "timepoint")) %>%
  full_join(cs_had_data, by = c("id", "period", "timepoint")) %>%
  full_join(hematological_data, by = c("id", "period", "timepoint")) %>%
  full_join(ferritin_data, by = c("id", "period", "timepoint")) %>% 
  full_join(blood_markers_data, by = c("id", "period", "timepoint")) %>% 
  
  mutate(bmi = weight/((height/100)^2), # weight in kg, height in meters
          
         vo2.max.bm = vo2.max/weight,
         vo2.max.lbm = vo2.max/lean.mass.whole.body,
    
         watt.max.bm = watt.max/weight,
         watt.max.lbm = watt.max/lean.mass.whole.body,
         
         watt.15tt.bm = watt.15tt/weight,
         watt.15tt.lbm = watt.15tt/lean.mass.whole.body,
         
         p.vo2.15tt = (vo2.15tt/vo2.max)*100,
         vo2.15tt.bm = vo2.15tt/weight,
         vo2.15tt.lbm = vo2.15tt/lean.mass.whole.body,
        
         watt.submax.bm = watt.submax/weight,
         watt.submax.lbm = watt.submax/lean.mass.whole.body,
         
         vo2.submax.bm = vo2.submax/weight,
         vo2.submax.lbm = vo2.submax/lean.mass.whole.body,
         
         hbmass.bm = hbmass/weight,
         hbmass.lbm = hbmass/lean.mass.whole.body,
         
         bv.bm = bv/weight,
         bv.lbm = bv/lean.mass.whole.body,
         
         rbcv.bm = rbcv/weight,
         rbcv.lbm = rbcv/lean.mass.whole.body,
         
         pv.bm = pv/weight,
         pv.lbm = pv/lean.mass.whole.body) %>%
  
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) %>% 
  # Filtering participants not completing training periods 1 and 2
  
  filter(!(id %in% c("101", "102", "103", "104", "105", "106", "107", "108", "109", "110", "111", "112", "114", "115", "116", "117"))) 
  # Filtering participants in other group
  
 

# Save as RDS for later use 
#saveRDS(test_data_wide, file = "data/derivedData/test_data_wide")


# Data in long format
test_data_long <- test_data_wide %>% 
  pivot_longer(
    cols = -c(id, period, timepoint, sex), # Exclude columns from pivoting
    names_to = "variable", 
    values_to = "value") %>% 
  group_by(id, variable) %>% 
  pivot_wider(names_from = timepoint,
              values_from = value) %>% 
  mutate(change.pre.post = post - pre) %>% 
  mutate(p.change.pre.post = ((post - pre)/pre) *100) %>% 
  group_by(variable) %>% print()


# Save as RDS for later use 
#saveRDS(test_data_long, file = "data/derivedData/test_data_long")


```

#### Questionnaire data

### Menopause data:

At study onset, female participants were asked whether they had reached menopause and, if so, the time elapsed since their last menstrual period.

```{r, menopause}
#| label: Menopause

menopause_data <- menopause.data %>% 
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) 
  # Filtering participants not completing training periods 1 and 2
  

# Save as RDS for later use 
#saveRDS(menopause_data, file = "data/derivedData/menopause_data")


```

### Physical activity level:

Participants were asked how many hh:mm per week they performed different types of activities. Here, the hh:mm is first converted to hours in decimals,

```{r, pa}
#| label: Physical activity


# Load data
activity_questionnaire_data <- fa_questionnaire.data %>% 
  
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) %>% 
  # Filtering participants not completing training periods 1 and 2
  
  filter(!(id %in% c("14", "27", "36", "40", "51"))) 
  # Filtering participants not answering the questionnaire for both period 1, the wash-out, or period 2
  

# Save as RDS for later use 
#saveRDS(activity_questionnaire_data, file = "data/derivedData/activity_questionnaire_data")


activity_before_intervention_questionnaire_data <- fa_questionnaire.data %>% 
  
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) %>% 
  filter(timepoint == "pre_intervention")

# Save as RDS for later use 
#saveRDS(activity_before_intervention_questionnaire_data, file = "data/derivedData/activity_before_intervention_questionnaire_data")

```

# Session data:

Data from all training sessions.

Participants completed \~ 24 sessions in each training period, alternating between 4x5-min all-out, 4x8-min high-intensity, and 6x6-min moderate-intensity.

```{r, sessions}
#| label: Session data 

# Finding pre-test values from period 1 and 2 to create fractions of maximal values

max.data <- test_data_wide %>% 
  select(id, period, timepoint, watt.max, hr.max) %>% 
  filter(timepoint == "pre") 


# Formatting the session data

session_data <- session.data %>% 
  filter(period %in% c(1, 2)) %>%
  mutate(id = as.factor(id)) %>% # Converting id to factor
  mutate(period = as.character(period)) %>% # Converting period to character
  
  filter(!(id == 15 & session.n == 22 & period == 1)) %>% # This was not a 4x5min session, plotting-mistake
  filter(!(id == 44 & session.n == 12 & period == 2)) %>%  # This was not a 4x5min session, plotting-mistake
  
  rename(session.type = type, 
         session.watt.4x5 = ftp,
         session.hr = hr,
         session.hr.max = hr.max, 
         session.watt = watt,
         session.rpe = rpe) %>%
  
  inner_join(max.data) %>%
  
  group_by(id, period) %>%
  arrange(session.n, .by_group = TRUE) %>%
  fill(session.watt.4x5, .direction = "down") %>% 
  ungroup() %>% 
  
  mutate(session.hr.max.percentage = (session.hr/hr.max)*100,
         session.watt.max.percentage = (session.watt/watt.max)*100,
         session.watt.4x5.percentage = (session.watt/session.watt.4x5)*100) %>% 
  
  group_by(id, period) %>%

  
  
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) 

  # Filtering participants not completing training periods 1 and 2
  
 
# Save as RDS for later use 
#saveRDS(session_data, file = "data/derivedData/session_data")


```

# VO2 session data

Data from the two 4x8-min sessions in period1 and 2 with VO2 measurements during work intervals

```{r, vo2 sessions}
#| label: VO2 session data

vo2_session_data <- vo2.session.data %>% 
  filter(period %in% c(1, 2)) %>%
  mutate(vo2.session.type = "4x8") %>% 
  rename(vo2.session.lac = lac, 
         vo2.session.rpe = rpe,
         vo2.session.watt = watt,
         vo2.session.vo2 = vo2,
         vo2.session.ve = ve,
         vo2.session.vco2 = vco2,
         vo2.session.bf = bf,
         vo2.session.hr = hr,
         vo2.session.rer = rer,
         vo2.session.vo2.percent = vo2.percent,
         vo2.session.hr.percent = hr.percent,
         vo2.session.90.vo2 = session.90.vo2,
         vo2.session.90.hr = session.90.hr) %>% 
  
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57")))  
  # Filtering participants not completing training periods 1 and 2
  

# Save as RDS for later use 
#saveRDS(vo2_session_data, file = "data/derivedData/vo2_session_data")


```

### CV calculation data from cycling tests

```{r, cv}
#| Label: CV

# CV calculation data (maximal and performance data)

cv_calculation_cycling_data <- filtered_max_data %>% 
  full_join(filtered_per_data, by = c("id", "group", "period", "timepoint")) %>%
  filter(timepoint %in% c("pre1", "pre2")) %>% 
  mutate(vo2.max.bm = vo2.max/weight,
         watt.max.bm = watt.max/weight,
         watt.15tt.bm = watt.15tt/weight) %>%
  select(-group) 

cv_calculation_cycling_data_p1_p2 <- cv_calculation_cycling_data %>% 
  filter(!(id %in% c("4", "6", "11", "12", "13" ,"20", "21", "25", "35", "37", "48", "52", "53", "54", "56", "57"))) 
  # Filtering participants not completing training periods 1 and 2
 

# Save as RDS for later use 
#saveRDS(cv_calculation_cycling_data_p1_p2, file = "data/derivedData/cv_calculation_cycling_data_p1_p2")


```
