---
title: "group-and-individual-training-responses-figure4"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---



```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data
## Test data
test_data_long <- readRDS(file = "data/derivedData/test_data_long")

test_data_wide <- readRDS(file = "data/derivedData/test_data_wide")


```


## Figure 4: Group and individual training responses

Creating a figure with 6 panels showing A-C) individual and group mean development in training outcomes in period 1 and 2, and D-F) individual and group mean change scores in period 2 and 2.


### Panels A-C: individual and group mean development in training outcomes in period 1 and 2

```{r}
#| label: Figure 4 A-C

# Saving sizes and widths of points and lines so easy to adjust
point.size = 0.7
line.width.individual = 0.3
line.width.mean = 0.5
errorbar.width.diff = 0.5
axis.width = 0.4
text.size = 9
title.text.size = 9



# Create a dummy plot to serve as empty space
empty <- ggplot() + theme_void()


# PO15-min

# Create plot showing mean and individual changes across training interventions
fig.timepoints <- test_data_wide %>%
  select(id, period, timepoint, watt.15tt) %>% 
  mutate(timepoint = ifelse(period == 1 & timepoint == "pre", "ET-1pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 1 & timepoint == "post", "ET-1post", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "pre", "ET-2pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "post", "ET-2post", timepoint)) %>%
  mutate(timepoint = factor(timepoint, levels = c("ET-1pre", "ET-1post",  "ET-2pre", "ET-2post"))) %>%  
  ggplot(., aes(x=timepoint, y=watt.15tt)) +
  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +
  geom_line(data = . %>% filter(timepoint %in% c("ET-1post", "ET-2pre")), 
            aes(group = 1), stat = "summary", fun = mean, 
            linetype = "dashed", color = "white", linewidth = line.width.mean) +
  scale_y_continuous(breaks = seq(from = 50, to = 300, by = 50)) +
  scale_x_discrete(
  labels = c(
    "ET-1pre"  = "pre",
    "ET-1post" = "post",
    "ET-2pre"  = "pre",
    "ET-2post" = "post"))+
  theme_pubr() + 
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size)) +
  labs(x = expression(" "),
       y = expression("MMPO"[15-min]*"  (W) ")) +
  
  # allow drawing outside plotting panel
    coord_cartesian(ylim = c(25, 300), clip = "off") +
    
   
    annotate("text", x = 1.5, y = -17, label = expression(ET[1]), size = 3.3) +
   annotate("text", x = 3.5, y = -17, label = expression(ET[2]), size = 3.3)


fig.timepoints


# Fit model and calculate 95% confidence intervals
model <- lmer(change.pre.post ~ period + (1|id), data = test_data_long %>% filter(variable == "watt.15tt"))
coefs <- broom.mixed::tidy(model, conf.int = TRUE) %>% print()


# Filter for period2 coefficient and create plot
fig.diff <- coefs %>%
  filter(term == "period2") %>%
  ggplot(aes(y = estimate, x = reorder(term, estimate))) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-4.06,ymax=4.06), fill="aliceblue") + # Adding colour to typical error = 4.06
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), linewidth = line.width.individual, width = errorbar.width.diff, color = "black") +
  geom_point(size = point.size, color = "black") +
  geom_hline(yintercept = 0, linetype = 2, linewidth = line.width.individual) +
  scale_y_continuous(position = "right", limits = c(-7, 17), breaks = seq(from = -5, to = 15, by = 5)) +
  theme_pubr() +
  labs(x = expression(""), 
       y = expression(~ Delta ~ " ET"[2] * "  vs. " ~ Delta ~  " ET"[1] * "  (W)"))+
  theme(text=element_text(size=text.size)) +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank(),  # Remove x-axis ticks
        axis.line.x = element_blank(),  # Remove x-axis line
        axis.title.x = element_blank(),  # Remove x-axis title
        legend.position = "none")
fig.diff


# Making the graph smaller
fig.diff <- plot_grid(empty, fig.diff, empty, 
            nrow = 3, rel_heights = c(0.75, 4, 1.25), 
            ncol = 1, align = "h")
fig.diff


# Adjust the plot margins
fig.timepoints <- fig.timepoints + 
  theme(plot.margin = margin(0.05, 0, 0.7, 0.05, "cm")) 

fig.diff <- fig.diff + 
  theme(plot.margin = margin(0.05, 0, 0, 0, "cm"))  

# Combine the plots
fig.watt.15tt.group <- plot_grid(fig.timepoints, fig.diff, 
            nrow = 1, rel_widths = c(4, 1))
fig.watt.15tt.group



# MPOvo2max

# Create plot showing mean and individual changes across training interventions
fig.timepoints <- test_data_wide %>%
  select(id, period, timepoint, watt.max) %>% 
  mutate(timepoint = ifelse(period == 1 & timepoint == "pre", "ET-1pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 1 & timepoint == "post", "ET-1post", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "pre", "ET-2pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "post", "ET-2post", timepoint)) %>%
  mutate(timepoint = factor(timepoint, levels = c("ET-1pre", "ET-1post",  "ET-2pre", "ET-2post"))) %>% 
  ggplot(., aes(x=timepoint, y=watt.max)) +
  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +
  geom_line(data = . %>% filter(timepoint %in% c("ET-1post", "ET-2pre")), 
            aes(group = 1), stat = "summary", fun = mean, 
            linetype = "dashed", color = "white", linewidth = line.width.mean) +
  scale_y_continuous(breaks = seq(from = 100, to = 400, by = 50)) +
  scale_x_discrete(
  labels = c(
    "ET-1pre"  = "pre",
    "ET-1post" = "post",
    "ET-2pre"  = "pre",
    "ET-2post" = "post"))+
  theme_pubr() + 
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size)) +
  labs(x = expression(" "), 
       y = expression("MMPO"[dot(V) * O[2 * max]] * " (W)")) +
  
  # allow drawing outside plotting panel
    coord_cartesian(ylim = c(80, 400), clip = "off") +
    
   
    annotate("text", x = 1.5, y = 30, label = expression(ET[1]), size = 3.3) +
   annotate("text", x = 3.5, y = 30, label = expression(ET[2]), size = 3.3)

fig.timepoints


# Fit model and calculate 95% confidence intervals
model <- lmer(change.pre.post ~ period + (1|id), data = test_data_long %>% filter(variable == "watt.max"))
coefs <- broom.mixed::tidy(model, conf.int = TRUE) %>% print()


# Filter for period2 coefficient and create plot showing difference in change score between periods
fig.diff <- coefs %>%
  filter(term == "period2") %>%
  ggplot(aes(y = estimate, x = reorder(term, estimate))) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-6.96,ymax=6.96), fill="aliceblue") + # Adding colure to typical error = 6.96
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), linewidth = line.width.individual, width = errorbar.width.diff, color = "#8b104e") + # Make red to show that grater change in ET-2 compared to ET-1
  geom_point(size = point.size, color = "#8b104e") +
  geom_hline(yintercept = 0, linetype = 2, linewidth = line.width.individual) +
  scale_y_continuous(position = "right", limits = c(-7, 17), breaks = seq(from = -5, to = 17, by = 5)) +
  theme_pubr() +
  labs(x = expression(""), 
       y = expression(~ Delta ~ " ET"[2] * "  vs. " ~ Delta ~  " ET"[1] * "  (W)"))+
  theme(text=element_text(size=text.size)) +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank(),  # Remove x-axis ticks
        axis.line.x = element_blank(),  # Remove x-axis line
        axis.title.x = element_blank(),  # Remove x-axis title
        legend.position = "none") 
  
fig.diff


# Making the graph smaller
fig.diff <- plot_grid(empty, fig.diff, empty, 
            nrow = 3, rel_heights = c(0.75, 4, 1.25), 
            ncol = 1, align = "h")
fig.diff



# Adjust the plot margins
fig.timepoints <- fig.timepoints + 
  theme(plot.margin = margin(0.05, 0, 0.7, 0.05, "cm")) 

fig.diff <- fig.diff + 
  theme(plot.margin = margin(0.05, 0, 0, 0, "cm"))  

# Combine the plots
fig.watt.max.group <- plot_grid(fig.timepoints, fig.diff, 
            nrow = 1, rel_widths = c(4, 1))
fig.watt.max.group



# VO2max figure

# Create plot showing mean and individual changes across training interventions
fig.timepoints <- test_data_wide %>%
  mutate(timepoint = ifelse(period == 1 & timepoint == "pre", "ET-1pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 1 & timepoint == "post", "ET-1post", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "pre", "ET-2pre", timepoint)) %>%
  mutate(timepoint = ifelse(period == 2 & timepoint == "post", "ET-2post", timepoint)) %>%
  mutate(timepoint = factor(timepoint, levels = c("ET-1pre", "ET-1post",  "ET-2pre", "ET-2post"))) %>% 
  ggplot(., aes(x=timepoint, y=vo2.max)) +
  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +
  geom_line(data = . %>% filter(timepoint %in% c("ET-1post", "ET-2pre")), 
            aes(group = 1), stat = "summary", fun = mean, 
            linetype = "dashed", color = "white", linewidth = line.width.mean) +
  scale_y_continuous(breaks = seq(from = 1000, to = 5000, by = 1000)) +
  scale_x_discrete(
  labels = c(
    "ET-1pre"  = "pre",
    "ET-1post" = "post",
    "ET-2pre"  = "pre",
    "ET-2post" = "post"))+
  theme_pubr() + 
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size)) +
  labs(x = expression(" "),
       y = expression(dot(V) * O[2 * max] * "  (mL ∙ min"^-1 * ") ")) +
  
  # allow drawing outside plotting panel
    coord_cartesian(ylim = c(600, 5000), clip = "off") +
    
   
    annotate("text", x = 1.5, y = -60, label = expression(ET[1]), size = 3.3) +
   annotate("text", x = 3.5, y = -60, label = expression(ET[2]), size = 3.3)
fig.timepoints


# Fit model and calculate 95% confidence intervals
model <- lmer(change.pre.post ~ period + (1|id), data = test_data_long %>% filter(variable == "vo2.max"))
coefs <- broom.mixed::tidy(model, conf.int = TRUE) %>% print()


# Filter for period2 coefficient and create plot
fig.diff <- coefs %>%
  filter(term == "period2") %>%
  ggplot(aes(y = estimate, x = reorder(term, estimate))) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-61.1,ymax=61.1), fill="aliceblue") + # Adding colure to typical error = 61.1 ml/min
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), linewidth = line.width.individual, width = errorbar.width.diff, color = "black") +
  geom_point(size = point.size, color = "black") +
  geom_hline(yintercept = 0, linetype = 2, linewidth = line.width.individual) +
  scale_y_continuous(position = "right", limits = c(-62, 110), breaks = seq(from = -50, to = 100, by = 50)) +
  theme_pubr() +
  labs(x = expression(""), 
       y = expression(~ Delta ~ " ET"[2] * "  vs. " ~ Delta ~  " ET"[1] * "  (mL ∙ min"^-1 * ") "))+
  theme(text=element_text(size=text.size)) +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank(),  # Remove x-axis ticks
        axis.line.x = element_blank(),  # Remove x-axis line
        axis.title.x = element_blank(),  # Remove x-axis title
        legend.position = "none")
fig.diff

# Making the graph smaller
fig.diff <- plot_grid(empty, fig.diff, empty, 
            nrow = 3, rel_heights = c(0.75, 4, 1.25), 
            ncol = 1, align = "h")
fig.diff


# Adjust the plot margins
fig.timepoints <- fig.timepoints + 
  theme(plot.margin = margin(0.05, 0, 0.7, 0.05, "cm")) 

fig.diff <- fig.diff + 
  theme(plot.margin = margin(0.05, 0, 0, 0, "cm"))  

# Combine the plots
fig.vo2.max.group <- plot_grid(fig.timepoints, fig.diff, 
            nrow = 1, rel_widths = c(4, 1))
fig.vo2.max.group




# Combining figures in long format
figure.4.a.b.c <- plot_grid(fig.watt.15tt.group, fig.watt.max.group, fig.vo2.max.group,
                       ncol = 1, rel_widths = c(1, 1, 1),
                       nrow = 3, rel_heights = c(1, 1, 1),
                       align = "v") +
  
  # Name panels
  draw_plot_label(x = c(0, 0, 0), 
                  y = c(0.99, 0.657, 0.325), 
                  label = c("A", "B", "C"), 
                  size = text.size) 

figure.4.a.b.c



```



### Panels D-F: individual and group mean change scores in period 2 and 2

```{r}
#| label: Figure 4 D-F

# Saving sizes and widths of points and lines so easy to adjust
point.size = 0.2
line.width = 0.3
line.width.individual = 0.3
line.width.mean = 0.5
axis.width = 0.4
text.size = 9


# Create a dummy plot to serve as empty space
empty <- ggplot() + theme_void()


# PO15min

fig.watt.15tt.change <- test_data_long %>%
  mutate(period = ifelse(period == 1, "ET-1", period)) %>%
  mutate(period = ifelse(period == 2, "ET-2", period)) %>%
  
  filter(variable == "watt.15tt") %>% 
  ggplot(aes(x=period, y=change.pre.post)) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-4.06,ymax=4.06), fill="aliceblue") + # Adding colure to typical error = 4.06
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = line.width.individual, colour = "black") +
  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  theme_pubr() +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +

  scale_y_continuous(limits = c(-20, 70), breaks = seq(from = -20, to = 70, by = 10)) +
  scale_x_discrete(
    labels = c(
    "ET-1"  = expression("ET"[1]),
    "ET-2" = expression("ET"[2]))) +
  
  labs(x = "Training period", y = expression(~ Delta ~ "MMPO"[15-min]*"  (W)")) +
  theme_pubr() +
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size),
        axis.title.x = element_blank(),
        legend.position = "none") 
fig.watt.15tt.change

fig.watt.15tt.change <- fig.watt.15tt.change + 
  theme(plot.margin = margin(0.05, 0, 0, 0.05, "cm"))



# POvomax

fig.watt.max.change <- test_data_long %>%  
  mutate(period = ifelse(period == 1, "ET-1", period)) %>%
  mutate(period = ifelse(period == 2, "ET-2", period)) %>%
  
  filter(variable == "watt.max") %>% 
  ggplot(aes(x=period, y=change.pre.post)) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-6.96,ymax=6.96), fill="aliceblue") + # Adding colure to typical error = 6.96
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = line.width.individual, colour = "black") +

  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +
  scale_y_continuous(limits = c(-20, 70), breaks = seq(from = -20, to = 70, by = 10)) +
  scale_x_discrete(
    labels = c(
    "ET-1"  = expression("ET"[1]),
    "ET-2" = expression("ET"[2]))) +
  
  labs(x = "Training period", y = expression(~ Delta ~ "MMPO"[dot(V) * O[2 * max]] * " (W)")) +
  theme_pubr() +
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size),
        axis.title.x = element_blank()) 
fig.watt.max.change

fig.watt.max.change <- fig.watt.max.change + 
  theme(plot.margin = margin(0.05, 0, 0, 0.05, "cm"))



# VO2max

fig.vo2.max.change <- test_data_long %>% 
  mutate(period = ifelse(period == 1, "ET-1", period)) %>%
  mutate(period = ifelse(period == 2, "ET-2", period)) %>%
  
  filter(variable == "vo2.max") %>% 
  ggplot(aes(x=period, y=change.pre.post)) +
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-61.1,ymax=61.1), fill="aliceblue") + # Adding colure to typical error = 61.1 ml/min
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = line.width.individual, colour = "black") +
  geom_line(aes(group = id), linewidth = line.width.individual, colour = "grey30", alpha = 0.3) +
  stat_summary(aes(colour="mean",group=1), fun = mean, geom="line", linewidth = line.width.mean, colour = "black") +
  scale_y_continuous(limits = c(-200, 700), breaks = seq(from = -200, to = 700, by = 100)) +
  
  scale_x_discrete(
    labels = c(
    "ET-1"  = expression("ET"[1]),
    "ET-2" = expression("ET"[2]))) +
  
  labs(x = "Training period", y = expression(~ Delta ~ dot(V) * O[2 * max] * "  (mL ∙ min"^-1 * ") ")) +
  theme_pubr() +
  theme(axis.line = element_line(linewidth = axis.width),
        text = element_text(size = text.size),
        axis.title.x = element_blank()) 
fig.vo2.max.change 

fig.vo2.max.change <- fig.vo2.max.change + 
  theme(plot.margin = margin(0.05, 0, 0, 0.05, "cm"))




# Figure 4 D-F:  MMPO15min, MPOvo2max, and VO2max
figure.4.d.e.f <- plot_grid(fig.watt.15tt.change, empty, fig.watt.max.change, empty, fig.vo2.max.change,
                       ncol = 1, rel_widths = c(1, 1, 1, 1, 1, 1),
                       nrow = 6, rel_heights = c(1, 0.095, 1, 0.095, 1, 0.095),
                       align = "v") +
  
  # Name panels
  draw_plot_label(x = c(0, 0, 0), 
                  y = c(0.99, 0.657, 0.325),
                  label = c("D", "E", "F"), 
                  size = text.size) 
figure.4.d.e.f


```


### Combining panels to one figure


```{r}
#| label: Figure 4 A-F

figure.4 <- plot_grid(figure.4.a.b.c, empty, figure.4.d.e.f,
                             ncol = 3, rel_widths = c(1, 0.05, 0.6), rel_heights = c(1, 1, 1))

figure.4

saveRDS(figure.4, "./figures/rds/figure.4.RDS")

ggsave("figures/figure.4.png", plot = figure.4, width =18, height = 28, 
       dpi = 800,
       units = "cm", device= png())


```




