---
title: "pre-test-analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data

## CV calculation data (for variables watt.15tt, watt.max, and vo2.max) 
cv_calculation_cycling_data_p1_p2 <- readRDS(file = "data/derivedData/cv_calculation_cycling_data_p1_p2")

## Test data
test_data_long <- readRDS(file = "data/derivedData/test_data_long")

```

## Pre-test analysis

1.  Test-retest reliability

2.  Reproducibility of individual pre-test values

3.  Effort during VO2max testing

### 1. Reliability of duplicate pre-tests

Percentage coefficient of variation (CV) of duplicate pre-tests were calculated using the guidelines provided by Hopkins (2000). Measures of Reliability in Sports Medicine and Science. Sports Med 30, 1–15 (2000). https://doi.org/10.2165/00007256-200030010-00001 Typical error of measurement (TE) is also calculated.

```{r}
#| label: Test-retest reliability

# Function to calculate CV and TE for multiple variables, grouping by id and period

calculate_reliability <- function(data, period_filter = NULL, variables) { 
  # Filter data by period, when a filter is provided
  data_filtered <- if (!is.null(period_filter)) {
    data %>% filter(period == period_filter)
  } else {
    data
  }
  
  # Loop over each variable in the "variables" vector
  results <- lapply(variables, function(var) {
    # Prepare the data for reliability calculations
    reliability_data <- data_filtered %>%
      select(id, period, timepoint, !!sym(var)) %>%
      group_by(id, period, timepoint) %>% # Group by id, period, and timepoint
      summarise(value = mean(!!sym(var), na.rm = TRUE), .groups = 'drop') %>% # Calculate mean by group
      pivot_wider(names_from = timepoint, values_from = value) %>% # Reshape data to wide format
      filter(complete.cases(pre1, pre2)) # Remove rows with NA values for pre1 and pre2
    
    # Calculate differences and reliability metrics
    reliability_metrics <- reliability_data %>%
      mutate(diff = pre2 - pre1) %>%  # Calculate the difference between pre2 and pre1
      summarise(
        s = sd(diff),                 # Standard deviation of the differences
        m = mean(rowMeans(cbind(pre1, pre2))), # Mean of pre1 and pre2
        te = s / sqrt(2),             # Typical error (TE)
        cv = 100 * (te / m)           # Percentage typical error (%TE)
      )
    

    
    reliability_metrics <- reliability_metrics %>%
      mutate(
        variable = var               # Add the variable name as a column
      )
    
    reliability_metrics
  })
  
  # Combine results for all variables into one data frame
  bind_rows(results)
}

# List of variables to calculate CV and TE for
variables <- c("vo2.max", "watt.max", "watt.15tt") # Only measures with double pre-tests

# Apply the function to calculate CV and TE for each period and both periods combined
reliability_p1 <- calculate_reliability(cv_calculation_cycling_data_p1_p2, period_filter = 1, variables = variables)  
# For period 1

reliability_p2 <- calculate_reliability(cv_calculation_cycling_data_p1_p2, period_filter = 2, variables = variables)  
# For period 2

reliability_combined <- calculate_reliability(cv_calculation_cycling_data_p1_p2, variables = variables)  
# For both period 1 and 2

# Print results
print(reliability_p1)       # Results for period 1
print(reliability_p2)       # Results for period 2
print(reliability_combined)  # Results for both period 1 and 2


```

### 2. Reproducibility of individual pre-test values

To assess the reproducibility of participants’ initial fitness level (pre-values), between ET1 and ET2, ICCs were estimated using linear mixed-effects models (lme4 package). For each outcome, a model was fitted on the form: outcome \~ 1 + (1 \| id) with participant id included as a random intercept to model between-individual differences in the outcome. In this model, the variance of the random intercept represents between individual variability, while the residual variance represents within-individual variability between training periods. ICCs were calculated as: ICC=σ_between/(σ_between+σ_within ) representing the proportion of total variance attributable to between-individual differences. Higher ICC estimates indicate greater consistency within individuals (i.e., low within-individual variability) and more stable differences between individuals (i.e., high between-individual variability). Bootstrapped 95% confidence intervals (CIs) for ICC estimates were computed using the bootMer function with 1000 resamples (lme4 package). The 95% CIs were interpreted as indicating poor (\< 0.50), moderate (0.50-0.75), good (0.75-0.90), or excellent (\>0.90) (Koo & Li, 2016).

Pearson´s correlation coefficients (r) with 95% CIs (R stats package) were used to examine associations between participants’ change scores, initial fitness level, recreational activity, and training load in ET1 and ET2.

```{r}
#| label: Reproducibility of individual pre-test values


# Outcomes to analyse

outcomes <- c("watt.max", "watt.15tt", "vo2.max", "p.vo2.15tt", "ge.submax", 
               "hbmass", "bv", "rbcv", "pv", "hct", "hb.c",
               "weight", "lean.mass.whole.body", 
               "ferritin.level",
               "fiber.area", "cd", "cf"
               ) 


## Intraclass correlation coefficients

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000, # 1000 resamples
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)       # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(pre))
  

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  
  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    pre ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )
summary(mod)
  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]             # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]       # Within-individual variance
    vb / (vb + vw)                            # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long,       # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf)



## Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
# Selecting pre values
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, pre) %>%
    filter(!is.na(pre)) %>% 
    pivot_wider(names_from = period, values_from = pre) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  
  
  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long, var)
}


# Format r results for presentation
r_summary <- results %>%
  mutate(
    `Pearsons r` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      pearsons_r,
      r_CI_lower,
      r_CI_upper
    )
  ) %>%
  select(variable, `Pearsons r`)  # Keep only relevant columns


# Print the results data frame
print(r_summary)


```



### 3. Effort during VO2max testing

```{r}
#| label: Effort during VO2max testing

# Preparing rer values from VO2max testing

  
# Mean (SD) rer values
mean_rer_values_data <- test_data_long %>% 
  filter(variable == "rer.max") %>% 
  select(id, period, pre, post) %>% 
  pivot_wider(
    names_from = period,
    values_from = c(pre, post),
    names_glue = "{.value}_p{period}") %>% 
  summarise(across(c(pre_p1, pre_p2, post_p1, post_p2),
                   list(mean_sd = ~sprintf("%.2f (%.2f)", mean(.x, na.rm = TRUE), sd(.x, na.rm = TRUE))),
                   .names = "{.col}")) %>% print()

rer_values_data <- test_data_wide %>% 
  select(id, period, timepoint, rer.max) %>% 
  mutate(variable = "rer.max") %>% 
  rename(value = rer.max) %>% print()
  

## Intraclass correlation coefficient
outcomes <- c("rer.max")


# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000, # 1000 resamples
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      timepoint = factor(timepoint),  # Timepoint as categorical fixed effect
      id     = factor(id)       # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(value))
  

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  
  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    value ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )
summary(mod)
  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]             # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]       # Within-individual variance
    vb / (vb + vw)                            # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = rer_values_data,       # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf)

```

