---
title: "training-responses-analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data
## Test data
test_data_long <- readRDS(file = "data/derivedData/test_data_long")

test_data_wide <- readRDS(file = "data/derivedData/test_data_wide")


```

## Training responses data

1.  Mean training responses in training periods 1 and 2, and difference in change

2.  Reproducibility of individual training responses

### 1. Mean training responses in training periods 1 and 2, and difference in change

Mean (SD) of pre, post, and change scores in training periods 1 and 2.

Potential differences in group-level training responses were investigated by fitting linear mixed models with change in the outcome variable of interest as the dependent variable, training period as the explanatory variable, and with participant-level intercepts (lme4 package).

```{r}
#| label: Mean training responses

## Mean (SD) of pre values for demographic table 

mean_values_pre_p1_data <- test_data_long %>% 
  filter(period == "1") %>% 
  select(id, period, sex, variable, pre) %>% 
  group_by(variable, sex) %>% 
  filter(variable %in% c("age", "height", "weight", "lean.mass.whole.body", "vo2.max", "vo2.max.bm", "watt.max", "watt.max.bm", "watt.15tt", "watt.15tt.bm")) %>% 
  summarise(across(c(pre),
                   list(mean_sd = ~sprintf("%.1f (%.1f)", mean(.x, na.rm = TRUE), sd(.x, na.rm = TRUE))),
                   .names = "{.col}")) %>%
  pivot_wider(names_from = sex,
              values_from = pre) %>% print()

## Mean (SD) at pre and post in both training interventions, as well as change scores to add in table

mean_values_timepoints_data <- test_data_long %>% 
  group_by(period, variable) %>% 

  filter(variable %in% c("age", "height", "weight", "lean.mass.whole.body", "vo2.max", "vo2.max.bm", "watt.max", "watt.15tt", "p.vo2.15tt", "ge.submax", "hbmass", "bv", "rbcv", "pv", "hct", "hb.c", "ferritin.level", "cs_umol.g.min", "had_umol.g.min", "fiber.area", "cd", "cf")) %>% 
  summarise(across(c(pre, post, change.pre.post, p.change.pre.post),
                   list(mean_sd = ~sprintf("%.1f (%.1f)", mean(.x, na.rm = TRUE), sd(.x, na.rm = TRUE))),
                   .names = "{.col}")) %>% 
  pivot_wider(names_from = period,
              values_from = c(pre, post, change.pre.post, p.change.pre.post))

flextable(mean_values_timepoints_data)



## Difference in changes between period 1 and 2

variables <- c("watt.max", "watt.15tt", "vo2.max", "vo2.max.bm", "p.vo2.15tt", "ge.submax", 
               "hbmass", "bv", "rbcv", "pv", "hct", "hb.c",
               "fiber.area", "cd", "cf",
               "weight", "lean.mass.whole.body") 

# Data frame to store results
results <- data.frame(
  variable = character(),
  estimate = numeric(),
  conf.low = numeric(),
  conf.high = numeric(),
  stringsAsFactors = FALSE
)

# Loop over each variable
for (var in variables) {
  cat("Processing variable:", var, "\n")
  
  # Fit model
  model <- lmer(change.pre.post ~ period + (1|id), data = test_data_long %>% filter(variable == var))
  print(model)
  
  # Calculate and print coefficients with confidence intervals
  coefs <- broom.mixed::tidy(model, conf.int = TRUE)
  print(coefs)
  
  # Extract only the second row (fixed effect for period)
  coefs_period <- coefs %>% filter(term == "period2")
  
  # Append results to the data frame
  results <- rbind(results, data.frame(
    variable = var,
    estimate = round(coefs_period$estimate, 2),
    conf.low = round(coefs_period$conf.low, 2),
    conf.high = round(coefs_period$conf.high, 2),
    stringsAsFactors = FALSE
  ))
}

# Print all results
print(results)
flextable(results)



## Difference in %-changes between ET-1 and ET-2

# List of variables to process
variables <- c("cs_umol.g.min", "had_umol.g.min")

# Data frame to store results
results <- data.frame(
  variable = character(),
  estimate = numeric(),
  conf.low = numeric(),
  conf.high = numeric(),
  stringsAsFactors = FALSE
)

# Loop over each variable
for (var in variables) {
  cat("Processing variable:", var, "\n")
  
  # Fit model
  model <- lmer(p.change.pre.post ~ period + (1|id), data = test_data_long %>% filter(variable == var))
  print(model)
  
  # Calculate and print coefficients with confidence intervals
  coefs <- broom.mixed::tidy(model, conf.int = TRUE)
  print(coefs)
  
  # Extract only the second row (fixed effect for period)
  coefs_period <- coefs %>% filter(term == "period2")
  
  # Append results to the data frame
  results <- rbind(results, data.frame(
    variable = var,
    estimate = round(coefs_period$estimate, 1),
    conf.low = round(coefs_period$conf.low, 1),
    conf.high = round(coefs_period$conf.high, 1),
    stringsAsFactors = FALSE
  ))
}


# Print all results
print(results)
flextable(results)


```

### 2. Reproducibility of individual training responses

```{r}
#| label: Reproducibility of individual training responses

## Absolute changes


# List of variables you want to process
outcomes <- c("watt.max", "watt.15tt", "vo2.max", "p.vo2.15tt", "ge.submax", 
               "hbmass", "bv", "rbcv", "pv", "hct", "hb.c",
               
               "watt.max.bm", "watt.max.lbm", "watt.15tt.bm", "watt.15tt.lbm",
               "vo2.max.bm", "vo2.max.lbm", 
               "hbmass.bm", "hbmass.lbm", "bv.bm", "bv.lbm",
               "rbcv.bm", "rbcv.lbm", "pv.bm", "pv.lbm",
               
               "fiber.area", "cd", "cf",
              
               "weight", "lean.mass.whole.body") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs

analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long,       # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, change.pre.post) %>%
    filter(!is.na(change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long, var)
}

# Print the results data frame
print(results)





## % changes of absolute values

# List of variables you want to process
outcomes <- c("watt.max", "watt.15tt", "vo2.max",
               "hbmass", "bv", "rbcv", "pv",
               
           
               "cs_umol.g.min", "had_umol.g.min",
               "fiber.area", "cd", "cf",
              
               "weight", "lean.mass.whole.body") 


# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(p.change.pre.post))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    p.change.pre.post ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = test_data_long,             # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf, n = Inf)





# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)

# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, p.change.pre.post) %>%
    filter(!is.na(p.change.pre.post)) %>% 
    pivot_wider(names_from = period, values_from = p.change.pre.post) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id) %>%
    subset(select = -variable)
  
  print(repeated_data)
  

  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(test_data_long, var)
}

# Print the results data frame
print(results)


```
