---
title: "training-characteristics-analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: Packages 

# Loading needed packages
source("./R/libs.R")


# Read the RDS files containing the data

## Exercise training data
session_data <- readRDS(file = "data/derivedData/session_data")

## Exercise training with VO2 measurements data
vo2_session_data <- readRDS(file = "data/derivedData/vo2_session_data")


```

## Training data

1.  Adherence to prescribed interval sessions

2.  Mean (sd) and individual reproducibility of training characteristics

3.  Reproducibility of individual MMPO 4x5-min response slopes



### 1. Adherence to prescribed interval sessions

```{r}
#| label: Training adherence

# Finding mean (SD) for training periods 1 and 2 to put in text
total_number_sessions <- session_data %>% 
  select(id, period, session.type) %>% 
  filter(!is.na(session.type)) %>%
  group_by(id, period) %>% 
  summarise(n = n()) %>% 
  group_by(period) %>% 
  summarise(n.m = mean(n, na.rm = TRUE),
            n.s = sd(n, na.rm = TRUE)) %>% print()

# Finding number of sessions performed in each period for each id
id_total_number_sessions <- session_data %>% 
  select(id, period, session.type) %>% 
  filter(!is.na(session.type)) %>%
  group_by(id, period) %>% 
  summarise(n = n()) %>% 
  mutate(n = as.numeric(n))%>%
  pivot_wider(names_from = "period",
              values_from = "n") %>% print()

# How many sessions each participant differed between period 1 and 2 were done manually


```

### 2. Mean and individual reproducibility of training characteristics

```{r}
#| label: Mean and individual reproducibility training


## Mean (SD) of training characteristics

# Finding mean (SD) power output, % of MMPOvo2max, HR, % of HRmax, and RPE for training periods 1 and 2
summarized.interval.session.data <- session_data %>%
  filter(!is.na(session.type)) %>%
  group_by(session.type, period) %>%
  summarise(across(c(session.rpe,
                     session.watt, session.watt.max.percentage, 
                     session.watt.4x5.percentage,
                     session.hr, session.hr.max.percentage),
                   list(mean_sd = ~sprintf("%.1f (%.1f)", mean(.x, na.rm = TRUE), sd(.x, na.rm = TRUE))),
                   .names = "{.col}"))

flextable(summarized.interval.session.data)


# Finding mean (SD) % of VO2max, time > 90% of VO2max, and lactate concentration during two 4x8-min interval sessions for training periods 1 and 2
vo2_session_mean_data <- vo2_session_data %>% 
  group_by(period) %>%
  summarise(across(c(vo2.session.vo2.percent, vo2.session.90.vo2, vo2.session.lac),
                   list(mean_sd = ~sprintf("%.1f (%.1f)", mean(.x, na.rm = TRUE), sd(.x, na.rm = TRUE))),
                   .names = "{.col}")) 

flextable(vo2_session_mean_data)



## Individual reproducibility of training characteristics

# 4x5-min, 4x8-min, and 6x6-min sessions in training periods 1 and 2

# Preparing data. Mean values grouped by type of session for each participant in each period
session_mean_data <- session_data %>% 
  filter(!is.na(session.type)) %>%
  group_by(id, period, session.type) %>% 
  
  summarise(mean.session.watt = mean(session.watt, na.rm = TRUE),
            mean.session.watt.max.percentage = mean(session.watt.max.percentage, na.rm = TRUE),
            
            mean.session.rpe = mean(session.rpe, na.rm = TRUE),
            
            mean.session.hr = mean(session.hr, na.rm = TRUE),
            mean.session.hr.max.percentage = mean(session.hr.max.percentage, na.rm = TRUE)) %>% 
  
  pivot_wider(names_from = session.type,
              values_from = mean.session.watt:mean.session.hr.max.percentage) %>% 
  pivot_longer(names_to = "variable",
               values_to = "value",
               cols = mean.session.watt_4x5:mean.session.hr.max.percentage_6x6) 



# List of variables you want to process
outcomes <- c("mean.session.watt_4x5", "mean.session.watt_4x8", "mean.session.watt_6x6",
               "mean.session.watt.max.percentage_4x5", "mean.session.watt.max.percentage_4x8", "mean.session.watt.max.percentage_6x6",
              
               "mean.session.rpe_4x5", "mean.session.rpe_4x8", "mean.session.rpe_6x6",
              
               "mean.session.hr_4x5", "mean.session.hr_4x8", "mean.session.hr_6x6",
               "mean.session.hr.max.percentage_4x5", "mean.session.hr.max.percentage_4x8", "mean.session.hr.max.percentage_6x6"
               )



# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(value))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    value ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]             # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]       # Within-individual variance
    vb / (vb + vw)                            # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = session_mean_data,    # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf)




# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)


# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, value) %>%
    filter(!is.na(value)) %>% 
    pivot_wider(names_from = period, values_from = value) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id)
  
  print(repeated_data)
  
  
  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  
  
  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(session_mean_data, var)
}

# Print the results data frame
print(results)





## Individual reproducibility of training characteristics

# 4x8-min sessions with vo2 measurements (two sessions in both training periods)


# Preparing data. Mean values grouped by type of session
vo2_session_mean_data <- vo2_session_data %>% 
  group_by(id, period) %>% 
  summarise(mean.vo2.session.lac = mean(vo2.session.lac, na.rm = TRUE),
            mean.vo2.session.vo2.percent = mean(vo2.session.vo2.percent, na.rm = TRUE),
            mean.vo2.session.90.vo2 = mean(vo2.session.90.vo2, na.rm = TRUE)) %>% 
  pivot_longer(names_to = "variable",
               values_to = "value",
               cols = c(mean.vo2.session.lac, mean.vo2.session.vo2.percent, mean.vo2.session.90.vo2)) 


# List of variables you want to process
outcomes <- c("mean.vo2.session.lac", "mean.vo2.session.vo2.percent", "mean.vo2.session.90.vo2")




# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(value))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    value ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))          # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = vo2_session_mean_data,    # Long-format dataset
  nboot = 1000                   # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf)




# Pearsons r

# Data frame to store results
results <- data.frame(
  variable = character(),
  r_CI_lower = numeric(),
  r_CI_upper = numeric(),
  pearsons_r = numeric(),
  stringsAsFactors = FALSE
)


# Function to process each variable
process_variable <- function(data, var_name) {
  repeated_data <- data %>%
    filter(variable == var_name) %>%
    select(id, period, variable, value) %>%
    filter(!is.na(value)) %>% 
    pivot_wider(names_from = period, values_from = value) %>%
    select(id, "1", "2") %>%
    rename(period.1 = `1`, period.2 = `2`) %>%
    subset(select = -id)
  
  print(repeated_data)
  
  
  # Pearsons r
  pearsons_r <- cor.test(repeated_data$period.1, repeated_data$period.2, method = "pearson")
  print(pearsons_r)
  

  # Extract r value and confidence intervals, and round to 2 decimals
  r_value <- round(pearsons_r$estimate, 2)
  r_ci_lower <- round(pearsons_r$conf.int[1], 2)
  r_ci_upper <- round(pearsons_r$conf.int[2], 2)
  
  # Append results to the data frame
  results <<- rbind(results, data.frame(
    variable = var_name,
    pearsons_r = r_value,
    r_CI_lower = r_ci_lower,
    r_CI_upper = r_ci_upper,
    stringsAsFactors = FALSE
  ))
}

# Applying the function to each variable
for (var in outcomes) {
  cat("Processing variable:", var, "\n")
  process_variable(vo2_session_mean_data, var)
}

# Print the results data frame
print(results)


```


### 3. Reproducibility of individual MMPO 4x5-min response slopes

```{r}
#| label: Reproducibility individual MMPO 4x5-min response slopes

# Watt at the five 4x5min sessions in period 1 and 2
session.watt.4x5_p1_p2 <- session_data %>%
  filter(session.type == "4x5") %>%
  mutate(session.n = min_rank(session.n)) %>%
  mutate(session.n = as.character(session.n)) %>%
  select(id, period, session.n, session.watt.4x5) %>% 
  
  # Some participants have missing values due values not noted in their training shemes (sessions were performed). To balance comparisons between periods, the value on the previous 4x5-min session is hence used.
  # Note that only 10 out of 420 values were missing.

  mutate(session.watt.4x5 = ifelse(period == 2 & id == 17 & session.n == 4, 269.5, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 27 & session.n == 4, 277.5, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 2 & session.n == 5, 107.5, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 30 & session.n == 4, 167.75, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 2 & id == 33 & session.n == 2, 91, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 36 & session.n == 4, 276, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 46 & session.n == 5, 180.75, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 2 & id == 49 & session.n == 5, 162.5, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 2 & id == 5 & session.n == 4, 160, session.watt.4x5)) %>%
  mutate(session.watt.4x5 = ifelse(period == 1 & id == 9 & session.n == 5, 228.75, session.watt.4x5))


# Calculating  each participant´s response slope in period 1 and 2 using the same linear mixed model as in above code chunck
model <- lmer(session.watt.4x5 ~ as.numeric(session.n)*period + 
                (as.numeric(session.n)*period | id),
              data = session.watt.4x5_p1_p2)


# Fixed effects (population-level estimates)
fixef_vals <- fixef(model)

# Random effects (individual deviations)
ranef_vals <- ranef(model)$id %>% 
  tibble::rownames_to_column("id") %>%
  rename(
    intercept = `(Intercept)`,
    slope_session.n = `as.numeric(session.n)`,
    slope_session.n_period2 = `as.numeric(session.n):period2`)

# Compute each participant’s full slope per period:
slope_data <- ranef_vals %>%
  
  mutate(
    slope.period1 = fixef_vals["as.numeric(session.n)"] + slope_session.n,
    slope.period2 = fixef_vals["as.numeric(session.n)"] +
                    fixef_vals["as.numeric(session.n):period2"] +
                    slope_session.n + slope_session.n_period2) %>%
  # slope.period1 = participant-specific slope in period 1 = population slope + participant deviation
  # slope.period2 = participant-specific slope in period 2 = population slope period 1 + population difference (period2 − period1) + participant deviation period 1 + participant deviation for period2 slope
  
  
  select(id, slope.period1, slope.period2) %>% 
  pivot_longer(
    cols = -id,
    names_to = "period",
    values_to = "value") %>%
  mutate(
    period = recode(period,
                    "slope.period1" = "1",
                    "slope.period2" = "2"),
    variable = "slope",
    value = value)



# Outcomes to analyse
outcomes <- c("slope") 



# Intraclass correlation

# Function to estimate ICCs with bootstrap CIs
analyze_trainability_icc_ci <- function(
  data,
  outcome,
  nboot = 1000,
  seed  = 123
) {

  message("Running model for: ", outcome)  # Progress message for tracking

  # Subset data to the current outcome and ensure correct factor structure
  df <- data %>%
    filter(variable == outcome) %>%
    mutate(
      period = factor(period),  # Training period as categorical fixed effect
      id     = factor(id)        # Participant ID as grouping factor
    ) %>% 
    filter(!is.na(value))

  # Skip analysis if too few observations are available
  if (nrow(df) < 5) {
    warning("Too few observations for ", outcome)
    return(NULL)
  }

  # Fit linear mixed-effects model to change scores
  # Fixed effect: training period (adjusts for systematic differences)
  # Random intercept: participant ID (captures between-individual variability)
  mod <- lmer(
    value ~ 1 + (1 | id),
    data = df,
    REML = TRUE  # REML provides less biased variance component estimates
  )

  # Function to extract ICC from a fitted mixed-effects model
  extract_icc <- function(fit) {
    vc <- as.data.frame(VarCorr(fit))         # Extract variance components
    vb <- vc$vcov[vc$grp == "id"]              # Between-individual variance
    vw <- vc$vcov[vc$grp == "Residual"]        # Within-individual variance
    vb / (vb + vw)                             # ICC calculation
  }

  # Compute point estimate of the ICC
  ICC_between <- extract_icc(mod)

  
  # Parametric bootstrap to obtain 95% CI for ICC
  boot_res <- tryCatch(
    bootMer(
      mod,
      FUN   = extract_icc,     # Statistic to bootstrap (ICC)
      nsim  = nboot,           # Number of bootstrap replicates
      use.u = TRUE,            # Preserve random-effects structure
      type  = "parametric",    # Parametric bootstrap
      seed  = seed
    ),
    error = function(e) NULL   # Gracefully handle bootstrap failures
  )

  # Extract percentile-based 95% confidence interval
  ci <- if (!is.null(boot_res)) {
    quantile(
      boot_res$t[, 1],
      probs = c(0.025, 0.975),
      na.rm = TRUE
    )
  } else {
    c(NA, NA)                  # Return NA if CI cannot be estimated
  }

  # Return tidy summary for the current outcome
  tibble(
    Variable         = outcome,
    ICC_between      = ICC_between,
    ICC_between_low  = ci[1],
    ICC_between_high = ci[2]
  )
}


# Run ICC analysis across all outcomes
results_icc <- map_dfr(
  outcomes,                     # Vector of outcome variable names
  analyze_trainability_icc_ci,  # Analysis function
  data  = slope_data,           # Long-format dataset
  nboot = 1000                  # Number of bootstrap replicates
)


# Format ICC results for presentation
icc_summary <- results_icc %>%
  mutate(
    `ICC Between` = sprintf(
      "%.2f [%.2f, %.2f]",      # Format ICC with 95% CI
      ICC_between,
      ICC_between_low,
      ICC_between_high
    )
  ) %>%
  select(Variable, `ICC Between`)  # Keep only relevant columns

# Print final table
print(icc_summary, width = Inf)




# Pearsons correlation

r_data <- slope_data %>%  
  filter(!is.na(value)) %>% 
  pivot_wider(names_from = period,
              values_from = value) %>%
  subset(select = -id) %>%
  print()


# Pearson r
pearsons.r <- r_data %>% 
  {cor.test(.$`1`, .$`2`, method = "pearson")} %>%
  print()



```



